---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rART)
library(ggplot2)
library(mlbench)
```

## ART is now available in R!

rART is a package containing the implementations of various ART (Adaptive Resonance Theory) models. ART is a class of neural network model that can perform both unsupervised and supervised (called ARTMAP) clustering online. The package is mainly written in Rcpp.

## What is Adaptive Resonance Theory?

Adaptive Resonance Theory (ART) is a neural network model developed by Stephen Grossberg. It is a neuro-cognitive model that attempts to explain how the brain learns and interacts with the stimuli in the environment. The main ART model consists of a winner-take-all system, with the ability to dynamically generate additional categories for capturing different patterns in the data^4^. These two properties help solve the stability-plasticity problem that plagues most machine learning algorithms: how to learn new patterns without erasing the learned memories.

ART also shows several advantages over other machine learning models. These include:

1. online learning, 
2. fast learning,
3. explanable.

For a comprehensive review of ART and its variants, please read da Silva et al., 2019 ^4^.

## The rART package

The rART package currently provides the implementation of two types of ART architecture: the standard ART and topology-learning ART (TopoART). TopoART forms linkages between clusters to enable the learning of the topology from the data. Additionally, TopoART possesses a noise filtering mechanism that prevents overfitting.

Currently, rART provides the fuzzy and hypersphere learning rules for both the standard and the topological versions. In the future, other learning rules and architectures will be added.

## Installation

rART is not available on CRAN yet, but you can install the latest development version:

```{r, eval = FALSE}
remotes::install_github("alersh/rART")
```

## Usage

### ART

ART is an unsupervised learning model and its object can be instantiated using the function ART(). Two learning rules are available: fuzzy and hypersphere. If the fuzzy rule is chosen, then all data must be normalized between 0 and 1^3^. Additionally, the complement of each data point d, which is 1 - d, must be created. Thus, if the dataset has a dimension of n, then the dimension of the actual input to the Fuzzy ART network is n x 2. Users are required to normalize the data prior to training and testing the data. Complement coding is taken care of by rART internally. The combination of the fuzzy rule and complement coding create hyperrectangular categories. 

The Fuzzy ART network has two hyperparameters: Learning rate ($\beta$) and vigilance ($\rho$)^3^. The learning rate $\beta$ must be between 0 and 1. Fast learning ($\beta$ = 1) can be achieved and is normally set as the default. The vigilance parameter $\rho$ decides whether an input pattern closely matches with the weight pattern of the selected category. The value must be between 0 and 1. A smaller value provides more generalization. 

Here is an example of using Fuzzy ART to cluster different shapes:

```{r, fig.width = 7, fig.height = 4}
trainShapes <- mlbench.shapes(n = 5000)
trainShapes$x <- normalize(trainShapes$x)
art <- ART(rule = "fuzzy", dimension = 2, vigilance = 0.93) # create an ART object
train(art, as.matrix(trainShapes$x)) 
plot(art, id = 0, .data = trainShapes$x) # plot the data and the weights.
```

### ARTMAP

ARTMAP is the supervised learning model^2^. rART provides both the standard and simplified versions of ARTMAP. For the summary descriptions on the standard and simplified ARTMAP, please see da Silva et al.^4^. For most classification problems, users should choose the simplified method as it is faster and uses less memory. The standard method is generally used for regression.

Here is a circle-in-a-square classification problem solved with the simplified Fuzzy ARTMAP:


```{r, fig.width = 7, fig.height = 4}
# circle in a square
trainCirSquare <- mlbench.circle(n = 10000)
testCirSquare <- mlbench.circle(n = 1000)
artmap <- ARTMAP(rule = "fuzzy", dimension = 2, vigilance = 0.8)
train(artmap, trainCirSquare$x, trainCirSquare$classes)
plot(artmap, .data = trainCirSquare$x, classes = trainCirSquare$classes) # create the 2 dimensional plot of the data and the weights
p <- predict(artmap, .data = testCirSquare$x, simpleTargets = testCirSquare$classes)
sum(p$matched)/length(p$matched) * 100 # percent correct
```

The user can also implement the standard fuzzy ARTMAP^2^. In this case, the user must first convert the class labels into dummy (binary) codes. rART provides two functions that aid the conversion. The createDummyCodeMap() takes the unique target labels and convert them to dummy codes. The function encodeLabel() converts all the target labels into dummy codes using this dummy code map. 

```{r, fig.width = 7, fig.height = 4}
# circle in a square
trainCirSquare <- mlbench.circle(n = 10000)
testCirSquare <- mlbench.circle(n = 1000)
artmap <- ARTMAP(rule = "fuzzy", dimension = 2, vigilance = 0.9, simplified = FALSE)
dummyMap <- createDummyCodeMap(unique(trainCirSquare$classes))
trainCirSquare$dummyClasses <- encodeLabel(trainCirSquare$classes, dummyMap)
testCirSquare$dummyClasses <- encodeLabel(testCirSquare$classes, dummyMap)
train(artmap, trainCirSquare$x, standardTargets = trainCirSquare$dummyClasses)
plot(artmap, .data = trainCirSquare$x, classes = trainCirSquare$dummyClasses, dummyCodeMap = dummyMap) # create the 2 dimensional plot of the data and the weights
p <- predict(artmap, .data = testCirSquare$x, standardTargets = testCirSquare$dummyClasses)
sum(p$matched, na.rm = T)/length(p$matched) * 100 # percent correct
```

The standard ARTMAP is better suited for regression problems^2^. Below is an example that uses the standard Fuzzy ARTMAP to approximate a sinusoidal function:

```{r}
x = seq(0, 1, by = 0.001)
y <- (sin(2*pi*x))^2
fn <- cbind(x = x, y = y) # function to be approximated
s <- sample(nrow(fn)) # randomize rows
s_train <- fn[s,]
a <- ARTMAP(dimension = 1, simplified = FALSE)
a$module$b$rho <- 0.95 # The degree of approximation is determined by the vigilance parameter in ART b.
train(a, as.matrix(s_train[,1]), standardTargets = as.matrix(s_train[,2]))
# get the fit
result <- predict(a, as.matrix(s_train[,1])) 
z <- cbind(s_train[,1], result$predicted)
ggplot(data.frame(x = z[,1],y = z[,2])) + geom_line(aes(x = x, y = y), color = "red") +
  geom_line(data = data.frame(x = s_train[,1], y = s_train[,2]), aes(x = x,y = y))

```

### Fuzzy TopoART

TopoART stands for "Topology-learning ART". It can learn the topology from the data by linking clusters together^5^. A TopoART network consists of two ART modules in series. Within each module, every learned category possesses a counter that counts the number of samples encoded by that category. A hyperparameter $\phi$ determines the minimum number of sample required for a category to become permanent. Every $\tau$ learning cycle, those categories containing samples fewer than $\phi$ will be removed. In the first module, those samples that belong to the permanent categories will be retained and advanced to the second module where they will be further classified into finer categories. Thus, the second module learns a subset of the data that are filtered by the first module. 

The learning of the topology can be done in both TopoART modules. During learning, once the best matching category is found, the second best matching category will be searched. If this second best mathcing category is found, then a link is created between this category and the best matching category. The learning rate for the second best matching category will be smaller than that for the best matching category.

Here is an example of using TopoART to cluster a noisy smiley face. Notice the smaller category size in module 2 when comparing it to the size in module 1. The user can specify which module to plot by entering the id number (0 for the first module and 1 for the second module).

```{r, fig.width = 7, fig.height = 4}
# smiley face
data(noisySmiley)
topoart <- TopoART(rule = "fuzzy", dimension = 2, vigilance = 0.88, tau = 200, phi = 5)
train(topoart, noisySmiley)
plot(topoart, id = 0, noisySmiley) # the ART a module
plot(topoart, id = 1, noisySmiley) # the ART b module
```

### Hypersphere ART

In all of the above examples, the user can substitute the fuzzy rule with the hypersphere rule^1^. The advantages of using the hypersphere rule are: 1) it does not require complement coding, and 2) it does not require normalization^1^.

## Final Words

I hope you will find this package useful. If you discover any problem, please contact me at alersh@gmail.com and provide me with the details of the problem. 



## References

1. Anagnostopoulos, GC, Georgiopoulos, M. (2000) "Hypersphere ART and ARTMAP unsupervised and supervised, incremental learning", Proceedings of the IEEE-INNS-ENNS International Joint Conference on Neural Networks. Neural Computing: New Challenges and Perspectives for the New Millennium, 6.

2. Carpenter, GA, Grossberg, S, Markuzon, N, Reynolds, JH, Rosen, DB. (1992) "Fuzzy ARTMAP: A neural network architecture for incremental supervised learning of analog multidimensional maps", IEEE Transactions on Neural Networks, 3(5), pp. 698-713.

3. Carpenter, GA, Grossberg, S, Rosen DB. (1991) "Fuzzy ART: Fast stable learning and categorization of analog patterns by an adaptive resonance system", Neural Networks, 4(6), pp 759-771.

4. da Silva, L.E.B, Elnabarawy, I., and Wunsch II, D.C. (2019) "A survey of Adaptive Resonance Theory neural network models for engineering applications", Neural Networks, 120, pp 167 - 203.

5. Tscherepanow, M. (2010) "TopoART: A topology learning hierarchical ART network", Proceedings of the International Conference on Artificial Neural Networks (ICANN). LNCS, 6354, pp. 157–167.
